Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-25 docs/manuscript.Rmd
updated

---
title: "Inferring the existence and direction of causal associations in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
bibliography: manuscript.bib
csl: nature.csl
---

@@

Measurement error will of course arise in other types of data too. For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and treating the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`). Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.

@@

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a matric to evaluate the sensitivity of the result of this extension to measurement error. Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$ (Model 2). For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:

@@

and a p-value is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical. The existence of causality and its direction is inferred based on the following scenarios:

@@

For the purposes of compiling simulation results, we use an arbitrary $\alpha = 0.05$ value.

@@

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages. First, it uses a formal statistical framework to test for the robustness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

@@

In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available. However, MR strategies exist for other scenarios, for example if there are multiple instruments for the hypothesised exposure then the problems described here may be circumvented [@DaveySmithHemani2014]. Likewise, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014]. We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].

In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

@@

>>>>

---
title: "Inferring the existence and direction of causal associations in the face of measurement error"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: no
    keep_tex: true
  word_document: default
bibliography: manuscript.bib
csl: vancouver.csl
---

@@

Measurement error will of course arise in other types of data too. For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data have been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

An established study design that can provide information about causality is randomisation. Given the hypothesis that trait A (henceforth referred to as the exposure) is causally related to trait B (henceforth referred to as the outcome), randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and treating the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This provides the theoretical basis behind randomised control trials, but in practice randomisation is often difficult or impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2003; @DaveySmithHemani2014]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation-based approaches and Mendelian randomisation (MR).

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`). Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must have its most proximal link with the exposure, associating with the outcome only through the exposure.

@@

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a metric to evaluate the sensitivity of the result of this extension to measurement error. Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT for each of two tests - $p_{CIT, x \rightarrow y}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$, $p_{CIT, y \rightarrow x}$ (Model 2). The results from each of these methods can then be used in combination to infer the existance and direction of causality. For some significance threshold $\alpha$ there are four possible outcomes from these two tests, and their interpretations are as follows:

@@

and a p-value, $p_{Steiger}$ is generated from the $Z$ value to indicate the probability of obtaining the observed difference in correlations $\rho_{gx}$ and $\rho_{gy}$ under the null hypothesis that both correlations are identical.

The existence of causality and its direction is inferred based on combining information from the MR analysis and the Steiger test. The MR analysis indicates whether there is a potential causal relationship ($p_{MR}$), and the Steiger test indicates the direction ($sign(Z)$) of the causal relationship and the confidence of the direction ($p_{Steiger}). For the purposes of compiling simulation results, these can be combined using an arbitrary $\alpha = 0.05$ value:

@@

@@

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing methylation levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages. First, it uses a formal statistical framework to test for the reliability of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

@@

In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available. We note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent. Multiple instruments can indeed help to circumvent some of the problems described here in other ways also, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014]. Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014]. We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c].

In this work we assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015; @Bowden2016b] have focused on accounting for the issues that horizontal pleiotropy can introduce when multiple instruments are available, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated. We have also not addressed other issues pertaining to instrumental variables, such as exposure missspecification, where the instrument is used for a particular variable due to that variable's non-causal correlation with the true exposure. However our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

@@

\raggedright
